---
title: "NYC Airbnb Price Analysis - Baseline Models"
author: "Andres Perez"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    highlight: tango
    code_folding: show
---

# Introduction

This document focuses on building baseline predictive models for Airbnb listing prices in New York City, using the cleaned dataset and engineered features from our EDA. We'll implement a stratified sampling approach to ensure proper representation of all boroughs in our training, testing, and evaluation sets.

# Required Libraries

```{r libraries}
library(dplyr)
library(stringr)
library(lubridate)
library(ggplot2)
library(caret)
library(RANN)
library(caTools)
library(tidyr)
library(purrr)
```

# Data Loading and Preparation

```{r load_data}
# Load the cleaned dataset
bnb = read.csv("/Users/andresperez/Desktop/R Files/bnb_project/data/bnb_listing_rev.csv")

# Remove ID column and duplicates
bnb = bnb[, !(names(bnb) == "id")]
bnb = distinct(bnb)

# Remove rows with zero price or NA price
bnb = bnb %>% filter(price > 0, !is.na(price))

# Process bathrooms
bnb = bnb %>%
  mutate(
    bathrooms = as.integer(sapply(strsplit(bathrooms_text, " "), "[", 1)),
    shared = sapply(strsplit(bathrooms_text, " "), function(x) {
      if(length(x) > 1) {
        ifelse(x[2] == "shared", 1, 0)
      } else {
        0
      }
    })
  )

# Process dates
bnb = bnb %>%
  mutate(
    last_review = mdy(last_review),
    last_review = as.integer(last_review),
    last_review_year = year(as.Date(last_review, origin = "1970-01-01")),
    host_since = mdy(host_since),
    host_since = as.integer(host_since),
    host_since_year = year(as.Date(host_since, origin = "1970-01-01"))
  )

# Convert categorical variables to factors
bnb = bnb %>%
  mutate(
    neighbourhood = as.factor(neighbourhood),
    neighbourhood_group = as.factor(neighbourhood_group),
    room_type = as.factor(room_type)
  )

# Add engineered features
bnb = bnb %>%
  mutate(
    price_per_person = price / accommodates,
    bed_efficiency = beds / accommodates,
    listing_age = 2024 - host_since_year,
    review_recency = 2024 - last_review_year,
    # Add room type indicators
    room_type_shared = as.numeric(room_type == "Shared room"),
    room_type_private = as.numeric(room_type == "Private room"),
    room_type_entire = as.numeric(room_type == "Entire home/apt"),
    # Add location-based features
    distance_from_center = sqrt((latitude - 40.7128)^2 + (longitude - (-74.0060))^2),  # Distance from NYC center
    # Add review-based features
    reviews_per_month = number_of_reviews / (2024 - host_since_year)  # Average reviews per month
  )

# Remove rows with NA values in key features
bnb = bnb %>%
  filter(!is.na(accommodates),
         !is.na(bedrooms),
         !is.na(beds),
         !is.na(bathrooms),
         !is.na(price_per_person),
         !is.na(bed_efficiency))

# Print number of rows after cleaning
cat("Number of rows after cleaning:", nrow(bnb), "\n")
```

# Data Splitting

```{r data_splitting}
# Set seed for reproducibility
set.seed(123)

# Create train/test/eval indices
n = nrow(bnb)
train_idx = sample(1:n, size = floor(0.6 * n))
test_idx = sample(setdiff(1:n, train_idx), size = floor(0.2 * n))
eval_idx = setdiff(1:n, c(train_idx, test_idx))

# Create splits
train_data = bnb[train_idx, ]
test_data = bnb[test_idx, ]
eval_data = bnb[eval_idx, ]

# Check distribution of boroughs in each split
cat("Training set borough distribution:\n")
print(table(train_data$neighbourhood_group))
cat("\nTesting set borough distribution:\n")
print(table(test_data$neighbourhood_group))
cat("\nEvaluation set borough distribution:\n")
print(table(eval_data$neighbourhood_group))
```

# Baseline Model Development

```{r baseline_model}
# Define features for modeling
features = c(
  # Original features
  "accommodates", "bedrooms", "beds", "bathrooms", "shared",
  "number_of_reviews", "last_review_year", "host_since_year",
  "latitude", "longitude",
  # Engineered features
  "price_per_person", "bed_efficiency", "listing_age", "review_recency",
  # Room type indicators ONLY (no 'room_type' factor)
  "room_type_shared", "room_type_private", "room_type_entire",
  # Location features
  "distance_from_center",
  # Review features
  "reviews_per_month"
)

# Train baseline model
formula = as.formula(paste("price ~", paste(features, collapse = " + ")))
baseline_model = lm(formula, data = train_data)

# Model summary
summary(baseline_model)

# Function to evaluate model
evaluate_model = function(model, data) {
  # Make predictions
  predictions = predict(model, newdata = data)
  
  # Remove any NA predictions
  valid_idx = !is.na(predictions)
  predictions = predictions[valid_idx]
  actual = data$price[valid_idx]
  
  # Calculate metrics
  rmse = sqrt(mean((actual - predictions)^2))
  mae = mean(abs(actual - predictions))
  r2 = 1 - sum((actual - predictions)^2) / sum((actual - mean(actual))^2)
  
  return(list(
    rmse = rmse,
    mae = mae,
    r2 = r2
  ))
}

# Evaluate on test set
test_metrics = evaluate_model(baseline_model, test_data)
cat("\nTest Set Metrics:\n")
cat("RMSE:", round(test_metrics$rmse, 2), "\n")
cat("MAE:", round(test_metrics$mae, 2), "\n")
cat("R-squared:", round(test_metrics$r2, 3), "\n")

# Evaluate on evaluation set
eval_metrics = evaluate_model(baseline_model, eval_data)
cat("\nEvaluation Set Metrics:\n")
cat("RMSE:", round(eval_metrics$rmse, 2), "\n")
cat("MAE:", round(eval_metrics$mae, 2), "\n")
cat("R-squared:", round(eval_metrics$r2, 3), "\n")
```

# Baseline Model Analysis

```{r model_analysis}
# Residual analysis
test_predictions = predict(baseline_model, newdata = test_data)
valid_idx = !is.na(test_predictions)
test_predictions = test_predictions[valid_idx]
test_residuals = test_data$price[valid_idx] - test_predictions

# Residual plot
ggplot(data.frame(
  predicted = test_predictions,
  residuals = test_residuals
), aes(x = predicted, y = residuals)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  labs(
    title = "Residual Plot",
    x = "Predicted Price",
    y = "Residuals"
  ) +
  theme_minimal()
```

*Figure: The residual plot shows the difference between actual and predicted prices for each listing. The funnel shape indicates increasing variance (heteroscedasticity) as predicted prices rise, suggesting the model is less accurate for higher-priced listings and may benefit from non-linear modeling or price transformation.*

```{r residual_distribution_plot}
# Residual distribution plot
ggplot(data.frame(residuals = test_residuals), aes(x = residuals)) +
  geom_histogram(bins = 50, fill = "skyblue") +
  labs(
    title = "Distribution of Residuals",
    x = "Residuals",
    y = "Count"
  ) +
  theme_minimal()
```

```{r feature_importance_plot}
# Feature importance plot
feature_importance = data.frame(
  feature = names(coef(baseline_model))[-1],  # Exclude intercept
  importance = abs(coef(baseline_model))[-1]
) %>%
  arrange(desc(importance))

# Filter out NA coefficients
feature_importance = feature_importance %>% filter(!is.na(importance))

ggplot(feature_importance, aes(x = reorder(feature, importance), y = importance)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  coord_flip() +
  labs(
    title = "Feature Importance",
    x = "Features",
    y = "Absolute Coefficient Value"
  ) +
  theme_minimal()
```

```{r summary_table, echo=FALSE}
library(knitr)
# Create a summary table for baseline model results
results_table <- data.frame(
  Set = c("Test", "Evaluation"),
  RMSE = c(round(test_metrics$rmse, 2), round(eval_metrics$rmse, 2)),
  MAE = c(round(test_metrics$mae, 2), round(eval_metrics$mae, 2)),
  R_squared = c(round(test_metrics$r2, 3), round(eval_metrics$r2, 3))
)
kable(results_table, caption = "Summary of Baseline Model Performance Metrics.")
```

## Next Steps

**Model Improvements:**
- Consider non-linear relationships
- Add more interaction terms
- Implement borough-specific models
- Try different algorithms

**Feature Engineering:**
- Create neighborhood clusters
- Add seasonal features
- Consider price transformations

**Model Selection:**
- Cross-validation
- Hyperparameter tuning
- Model ensemble approaches
